Dependently Typed Languages in Statix

### 2023-03-15 pres-evcs/presentation.tex
Presentation feedback

\begin{frame}[fragile]{Background: What are Dependent Types?}
\begin{itemize}
	\item Types may depend on values!
	\begin{exampleblock}{Example}
		\texttt{concat : (A: Set) -> (n : Nat) -> Vec A n -> Vec A n
			\\ \hspace*{48pt} -> Vec A (n + n)}
	\end{exampleblock}
% - n is a RUNTIME value, it does not have to be known at compile time
% - n occurs insie a type `Vec`!
% - The type system enforces these lengths are always correct
% - Challenge: Type checking may require evaluating arbitrary terms to decide equality
	\item Curry-Howard correspondence
% Can be used to write proofs about code
\end{itemize}

@@

\begin{frame}[fragile]{Research Question}
How well Statix is fit for the task of defining a dependently-typed language.
% Statix is turing complete, so it's going to be possible
% But is it going to be easier or harder than a general purpose language like haskell?
\end{frame}

\begin{frame}[fragile]{Why is this important?}
	\begin{block}{From the perspective of Spoofax research}
		Developing a language with a complex type system tests the boundaries of what Spoofax can do.
	\end{block}
	% - What can we improve about Statix?

	\begin{block}{From the perspective of Dependent Types research}
		A rapid prototyping platform.
	\end{block}
	% - Try a new feature without having to implement in general purpose
	% - Spoofax provides an easy way to get a parser, type checker, etc
\end{frame}

@@

\begin{frame}[fragile]{Type Checking}
	\begin{block}{Type checking relation}
		\texttt{typeOfExpr : scope * Expr -> Expr}
	\end{block}
	% Returns an Expr, not a Type
	% Scope -> Two different types of nodes
	\begin{block}{How do we use scopes?}
		One relation \texttt{name $\rightarrow$ NameEntry}, NameEntry is either:
		\begin{itemize}
			\item NType: Stores a type
			\item NSubst: Stores a substitution
		\end{itemize}
	\end{block}
\end{frame}

\begin{frame}[fragile]{Type Checking: Requires Evaluation}
	\begin{exampleblock}{Example 1}
		\texttt{let T = Bool;\\let b: T = true;}
	\end{exampleblock}
	\begin{block}{Evaluation relation} 
		\texttt{betaHeadReduce : scope * Expr -> scope * Expr} \\
		\texttt{betaReduce : scope * Expr -> Expr} \\
		\texttt{exectBetaEq : (scope * Expr) * (scope * Expr)}
	\end{block}
\end{frame}

\begin{frame}[fragile]{Type Checking: Rules}
	\begin{figure}
		\centering
		\includegraphics[width=0.5\linewidth]{img/screenshot001}
		\label{fig:screenshot001}
	\end{figure}
	\begin{figure}
		\centering
		\includegraphics[width=0.5\linewidth]{img/screenshot002}
		\label{fig:screenshot002}
	\end{figure}

@@

\begin{frame}[fragile]{Extra contributions}
\begin{enumerate}
	\item Implemented Inference
	\item Implemented Inductive Data Types
	\item Implemented Universes
	\item Interpreter
	\item Compiler to Clojure
	\item Comparison with implementation in Haskell
	\item Comparison with implementation in LambdaPi
	\item Evaluation of Spoofax
\end{enumerate}
\end{frame}

\begin{frame}[fragile]{Conclusions}
	Spoofax is a great tool for developing dependently typed languages!\footnote{But there is still room for improvement}
\end{frame}

>>>>

\begin{frame}[fragile]{Background: What are Dependent Types?}
\begin{itemize}
	\item Types may depend on values!
	\begin{exampleblock}{Example}
		\texttt{concat : (A: Set) -> (n m : Nat) -> Vec A n -> Vec A m
			\\ \hspace*{48pt} -> Vec A (n + m)}
	\end{exampleblock}
% - n is a RUNTIME value, it does not have to be known at compile time
% - n occurs insie a type `Vec`!
% - The type system enforces these lengths are always correct
% - Challenge: Type checking may require evaluating arbitrary terms to decide equality
	\item Proof assistants
	\item For example: Agda, Coq, Lean, ...
% Can be used to write proofs about code
\end{itemize}

@@

\begin{frame}[fragile]{Research Question}
\large{How suitable is Statix for defining a dependently-typed language?}
% Statix is turing complete, so it's going to be possible
% But is it going to be easier or harder than a general purpose language like haskell?
\end{frame}

\begin{frame}[fragile]{Why is this important?}
	\begin{block}{Spoofax perspective}
		Developing a language with a complex type system tests the boundaries of what Spoofax can do.
	\end{block}
	% - What can we improve about Statix?

	\begin{block}{Dependent Types perspective}
		Using a language workbench helps with rapid prototyping.
	\end{block}
	% - Try a new feature without having to implement in general purpose
	% - Spoofax provides an easy way to get a parser, type checker, etc
\end{frame}

\begin{frame}{Primary Contribution}
	todododododo
\end{frame}

@@

\begin{frame}[fragile]{Type Checking}
	\begin{block}{Type checking relation}
		\texttt{typeOfExpr : scope * Expr -> Expr}
	\end{block}
	% Returns an Expr, not a Type
	% Scope -> Two different types of nodes
	\begin{block}{How do we use scopes?}
		A scope is used as a combination of an environment and a context.\\
		One relation \texttt{name $\rightarrow$ NameEntry}, NameEntry is either:
		\begin{itemize}
			\item NType: Stores a type -> Corresponds with context
			\item NSubst: Stores a substitution -> Corresponds with environment
		\end{itemize}
	\end{block}
\end{frame}

\begin{frame}[fragile]{Type Checking: Requires Evaluation}
	\begin{exampleblock}{Example 1}
		\texttt{let T = if false then Int else Bool end;\\let b: T = true;}
	\end{exampleblock}
	\begin{block}{Evaluation relation} 
		\texttt{betaHeadReduce : scope * Expr -> scope * Expr} \\
		\texttt{betaReduce : scope * Expr -> Expr} \\
		\texttt{exectBetaEq : (scope * Expr) * (scope * Expr)}
	\end{block}
\end{frame}

\begin{frame}[fragile]{Type Checking: From inference rules to Statix code}
	todododododod From inference rules to Statix code

	\begin{figure}
		\centering
		\includegraphics[width=0.5\linewidth]{img/screenshot001}
		\label{fig:screenshot001}
	\end{figure}
	\begin{figure}
		\centering
		\includegraphics[width=0.5\linewidth]{img/screenshot002}
		\label{fig:screenshot002}
	\end{figure}

@@

\begin{frame}[fragile]{Extra contributions}
	% Refer to thesis for this
Features
\begin{enumerate}
	\item Implemented Inference
	\item Implemented Inductive Data Types
	\item Implemented Universes
	\item Interpreter
	\item Compiler to Clojure
\end{enumerate}
Evaluation
\begin{enumerate}
	\item Comparison with implementation in Haskell
	\item Comparison with implementation in LambdaPi
	\item Evaluation of Spoofax
\end{enumerate}
\end{frame}

\begin{frame}[fragile]{Conclusions}
	Spoofax is a great tool for developing dependently typed languages!\footnote{But there is still room for improvement}

	- using scopes as env+context 
	todo beperkingen noemen 
	- statix enforces syntactic uniqueness of metavariable solutions but dependently typed languages use a weaker notion of equality 
	- full thesis coming soonTM
\end{frame}
